### 2011-08-11 paper_scratch/abstract.tex
Added a scratch version to produced reordered table for ECAL presentation

Creating gaits for legged robots is an important task to enable robots
to access rugged terrain, yet designing such gaits by hand is a
challenging and time-consuming process. In this paper we investigate
various algorithms for automating the creation of quadruped
gaits. Because many robots do not have accurate simulators, we test
gait learning algorithms entirely on a physical robot.  We compare the
performance of two classes of learning gaits: locally searching
parameterized motion models and evolving artificial neural networks
with the HyperNEAT generative encoding. Specifically, we test six
different parameterized learning strategies: uniform and Gaussian
random hill climbing, policy gradient reinforcement learning,
Nelder-Mead simplex, a random baseline, and a new method that builds a
model of the fitness landscape with linear regression to guide further
exploration.  While all parameter search methods outperformed a
manually designed gait, only the linear regression and Nelder-Mead
simplex strategies outperform a random baseline strategy. Gaits
evolved by HyperNEAT perform considerably better than all
parameterized local search methods and produced gaits nearly 9 times faster than a hand-designed gait.  The best evolved HyperNEAT gaits exhibit
complex motion patterns that contain multiple frequencies, yet are
regular in that the leg movements are coordinated.

>>>>

Creating gaits for legged robots is an important task to enable robots
to access rugged terrain, yet designing such gaits by hand is a
challenging and time-consuming process. In this paper we investigate
various algorithms for automating the creation of quadruped
gaits. Because many robots do not have accurate simulators, we test
gait-learning algorithms entirely on a physical robot.  We compare the
performance of two classes of gait-learning algorithms: locally searching
parameterized motion models and evolving artificial neural networks
with the HyperNEAT generative encoding. Specifically, we test six
different parameterized learning strategies: uniform and Gaussian
random hill climbing, policy gradient reinforcement learning,
Nelder-Mead simplex, a random baseline, and a new method that builds a
model of the fitness landscape with linear regression to guide further
exploration.  While all parameter search methods outperform a
manually-designed gait, only the linear regression and Nelder-Mead
simplex strategies outperform a random baseline strategy. Gaits
evolved with HyperNEAT perform considerably better than all
parameterized local search methods and produce gaits nearly 9 times faster than a hand-designed gait.  The best HyperNEAT gaits exhibit
complex motion patterns that contain multiple frequencies, yet are
regular in that the leg movements are coordinated.

### 2011-08-11 paper_scratch/introduction.tex
Added a scratch version to produced reordered table for ECAL presentation

An additional motivation is to test whether gaits evolved in simulation transfer to reality well, especially those evolved with cutting-edge evolutionary algorithms. Because HyperNEAT gaits performed well in
simulation, it is interesting to test whether HyperNEAT can produce fast
gaits for a physical robot, including handling the noisy, unforgiving nature of the real world. Such tests help us better understand the real world implications of results reported only in simulation. It is additionally interesting to test how
more traditional gait optimization techniques compete with
evolutionary algorithms when evolving in hardware. A final motivation of this research is simply to evolve effective gaits for a physical robot.

>>>>

An additional motivation is to test whether techniques for evolving
gaits in simulation, especially cutting-edge evolutionary algorithms,
transfer to reality well. Because HyperNEAT gaits performed well in
simulation, it is interesting to test whether HyperNEAT can produce fast
gaits for a physical robot, including handling the noisy, unforgiving nature of the real world. Such tests help us better understand the real world implications of results reported only in simulation. It is additionally interesting to test how
more traditional gait optimization techniques compete with
evolutionary algorithms when evolving in hardware. A final motivation of this research is simply to evolve effective gaits for a physical robot.

### 2011-08-11 paper_scratch/results.tex
Added a scratch version to produced reordered table for ECAL presentation

%%%%%%% NEW TABLE
\begin{table}
\begin{center}
\begin{tabular}{|r|c|c|c||c|}
\hline
                                         & Average & Std. Dev. \\
\hline                                    
\hline                                    
Previous hand-coded gait                 & 5.16   &   --     \\
\hline
Random search                            & 9.40   &   6.83   \\
\hline
Uniform Random Hill Climbing             & 7.83   &   4.56   \\
\hline
Gaussian Random Hill Climbing            & 10.03  &   6.00   \\
\hline
Policy Gradient Descent                  & 6.32   &   7.39   \\
\hline
Nelder-Mead simplex                      & 12.32  &   3.35   \\
\hline
Linear Regression                        & 14.01  &  12.88   \\
\hline
Evolved Neural Network              &        &          \\
(HyperNEAT)                              & 29.26  &   6.37   \\
\hline
\end{tabular}
\caption{The average and standard deviation of the best gaits found
  for each algorithm during each of three runs, in body lengths per
  minute.}  \tablabel{results}
\end{center}
\end{table}

@@

The results for the gaits evolved by HyperNEAT are shown in
\figref{hnResults} and \tabref{results}.  A total of 540 evaluations
were performed for HyperNEAT (180 in each of three runs).
Overall the HyperNEAT gaits
were the fastest by far, beating all the parameterized models when
comparing either average or best gaits.  We believe that this is
because HyperNEAT was allowed to explore a much richer space of
motions, but did so while still utilizing symmetries when
advantageous.  The single best gait found during this study had a
speed of 45.72 body lengths / minute, 66\% better than the best
non-HyperNEAT gait and 8.9 times faster than the hand-coded gait.
\figref{neat_110115_211410_00000_002_filt_zoom} shows a typical
HyperNEAT gait that had high fitness.  The pattern of motion is both
complex (containing multiple frequencies and repeating patterns across
time) and regular, in that patterns of multiple motors are
coordinated.

The evaluation of the gaits produced by HyperNEAT was more noisy than
for the parameterized gaits, which made learning difficult. For
example, we tested an example HyperNEAT generation-champion gait 11
times and found that its mean performance was 26 body lengths per
minute ($\pm$ 13 SD), but it had a max of 38 and a min of 3.  Many
effective HyperNEAT gaits were not preserved across generations
because a single poor-performing trial could prevent their selection. The HyperNEAT learning curve would be smoother if
the noise in the evaluations could be reduced or more than one
evaluation per individual could be afforded.

>>>>

%%%%%%% NEW TABLE
\begin{table}
\begin{center}
\begin{tabular}{|r|c|c|c||c|}
\hline
                                         & Average & Std. Dev. \\
\hline                                    
\hline                                    
Previous hand-coded gait                 & 5.16   &   --     \\
\hline
Random search                            & 9.40   &   6.83   \\
\hline
Uniform Random Hill Climbing             & 7.83   &   4.56   \\
\hline
Gaussian Random Hill Climbing            & 10.03  &   6.00   \\
\hline
Policy Gradient Descent                  & 6.32   &   7.39   \\
\hline
Nelder-Mead simplex                      & 12.32  &   3.35   \\
\hline
Linear Regression                        & 14.01  &  12.88   \\
\hline
Evolved Neural Network              &        &          \\
(HyperNEAT)                              & 29.26  &   6.37   \\
\hline
\end{tabular}
\caption{The average and standard deviation of the best gaits found
  for each algorithm during each of three runs, in body
  lengths/minute.}  \tablabel{results}
\end{center}
\end{table}

@@

The results for the gaits evolved by HyperNEAT are shown in
\figref{hnResults} and \tabref{results}.  A total of 540 evaluations
were performed for HyperNEAT (180 in each of three runs).
Overall the HyperNEAT gaits
were the fastest by far, beating all the parameterized models when
comparing either average or best gaits.  We believe that this is
because HyperNEAT was allowed to explore a much richer space of
motions, but did so while still utilizing symmetries when
advantageous.  The single best gait found during this study had a
speed of 45.72 body lengths/minute, 66\% better than the best
non-HyperNEAT gait and 8.9 times faster than the hand-coded gait.
\figref{neat_110115_211410_00000_002_filt_zoom} shows a typical
HyperNEAT gait that had high fitness.  The pattern of motion is both
complex (containing multiple frequencies and repeating patterns across
time) and regular, in that patterns of multiple motors are
coordinated.

The evaluation of the gaits produced by HyperNEAT was more noisy than
for the parameterized gaits, which made learning difficult. For
example, we tested an example HyperNEAT generation-champion gait 11
times and found that its mean performance was 26 body lengths/minute
($\pm$ 13 SD), but it had a max of 38 and a min of 3.  Many
effective HyperNEAT gaits were not preserved across generations
because a single poor-performing trial could prevent their selection. The HyperNEAT learning curve would be smoother if
the noise in the evaluations could be reduced or more than one
evaluation per individual could be afforded.
